Road User video evidence of road traffic offences: Preliminary analysis of Operation Snap data and suggestions for a research agenda

### 2024-07-25 paper.qmd
Typo fixes

There were `r nrow(d_all) |> scales::number(big.mark = ",")` records in the dataset for the three-year study period, with a strong upward trend, as shown in the monthly counts presented in @fig-time.
Since early 2022, there have been more monthly records in the OpSnap data than in the official 'STATS19' road traffic collision records for West Yorkshire, highlighting the under-reporting of road traffic incidents in official statistics.
STATS19 records are from the Department for Transport's database of road traffic collisions reported to the police and only include incidents that result in injury.
Like OpSnap data, STATS19 records are open access.
They were downloaded with the `stats19` R package [@lovelace2019] and filtered to include only records from West Yorkshire to ensure comparability with the OpSnap data.

@@

```{r}
#| label: tbl-offences
#| include: true
#| tbl-cap: "Number of convictions by offence type (top 6 and other)."
#| width: 60%
# Aim: get table of n. offences by mode
d_mode_offence_count = d_all |>
  count(mode, offence, sort = TRUE) 
# offences in order of n. offences
d_offence_count = d_all |>
  count(offence, sort = TRUE)
# d_offence_count |>
#   arrange(desc(n)) |>
#   head(20) |>
#   knitr::kable()
# |offence                                                                                                |    n|
# |:------------------------------------------------------------------------------------------------------|----:|
# |n/a                                                                                                    | 5706|
# |rt88576 drive without reasonable consideration to others                                               | 4992|
# |rt88575 drive without due care and attention                                                           | 2917|
# |rt88975 drive motor vehicle fail to comply with red / green arrow / lane closure traffic light signals | 1364|
# |rt88971 fail to comply with red traffic light                                                          |  679|
# |rt88966 motor vehicle fail to comply with endorsable s36 traffic sign                                  |  411|
# |rv86019 use a handheld phone / device whilst driving a motor vehicle on a road                         |  357|
# |rt88760 fail to comply with solid white lines                                                          |  265|
# |rt88751 contravene give way sign                                                                       |  264|
# |suspected contravene weight restriction.                                                               |  213|
# |rt88751 contravene mandatory direction arrows                                                          |  212|
# |me82009 driving on hard shoulder of motorway                                                           |  113|
# |rt88975 fail to comply with red traffic light                                                          |  109|
# |rt88751 motor vehicle fail to comply with a non-endorsable traffic sign other (specify)                |   91|
# |zp97004 fail to comply with red light pelican crossing                                                 |   84|
# |zp97003 stop within controlled area of pelican crossing                                                |   81|
# |rc86814 driver not in proper control of vehicle                                                        |   62|
# |zp97001 stop vehicle within limits of pelican crossing                                                 |   59|
# |hy35001 drive/ride on footpath beside a road                                                           |   42|
# |rr84171 vehicle contravene local traffic order other than parking (e.g. bus lane)                      |   41|
# Pull out the top 6 offences excluding n/a:
d_offence_top_6 = d_mode_offence_count |>
  filter(offence != "n/a") |>
  group_by(offence) |>
  summarise(n = sum(n)) |>
  arrange(desc(n)) |>
  head(6)
d_offence_classified = d_offence_count |>
  filter(offence != "n/a") |>
  mutate(
    Offence = case_when(
      offence %in% d_offence_top_6$offence ~ offence,
      TRUE ~ "Other"
    )
  ) |>
  group_by(Offence) |>
  summarise(`Number of records` = sum(n)) |>
  # Arrange in descending order of n except for "Other":
  arrange(Offence == "Other", desc(`Number of records`)) |>
  mutate(`Percent of records` = round(`Number of records` / sum(`Number of records`) * 100, 1))

@@

tbl = d_offence_classified |>
  mutate(
    `Percent of records`= round(`Percent of records`)
    ) |> 
  rename(`Number` = `Number of records`, `Percent`=`Percent of records`) |> 
  gt() |>

  gt_plt_bar_pct(column = `Percent`, fill = "#252525", scaled = TRUE, labels=TRUE, width=120, height=30, font_size = "15px") |> 
  text_transform(
    fn <- function(x){
      code <- str_extract(x, "^[^\ ]+")
      desc <- str_remove(x, "^[^\ ]+")
      glue::glue("<em><span style='font-size:14px'>{code}</span></em><br><span style='font-size:18px'>{desc}</span>")
      },
    locations=cells_body(columns=Offence)
  ) |> 
  cols_width(1 ~ px(600), 2 ~ px(80)) |> 
  gt_theme_espn()
# gt::gtsave(tbl, "tbl-offences.png")
# browseURL("tbl-offences.png")
knitr::include_graphics("tbl-offences.png")
```

```{r}
# reclassify offences into types
x = d_all
regroup_offences_simple = function(x) {
 x |> 
   dplyr::mutate(
     offence_simple = dplyr::case_when(
       stringr::str_detect(offence, "drive without reasonable consideration") ~ "Inconsiderate driving",
       stringr::str_detect(offence, "drive without due care") ~ "Careless driving",
       TRUE ~ "Other"

     )
   )
}
# table(x$mode)
regroup_modes = function(x) {
  x |> 
    dplyr::mutate(
      mode_simplified = dplyr::case_when(
      mode == "cyclist" ~ "Cyclist",
      mode == "vehicle driver" ~ "Driver",
      TRUE ~ "Other"
    ))
}

@@

```{r}
#| label: tbl-mode
#| tbl-cap: Number of OpSnap records by observer mode.
#| width: "80%"
# d_all |>
#   count(mode, sort = TRUE) |>
#   mutate(percent_of_records = n / nrow(d_all)) |>
#   mutate(percent_of_records = round(percent_of_records, 3) * 100) |>
#   arrange(desc(n)) |>
#   rename_all(snakecase::to_title_case) |>
#   # Rename N to "Number of records"
#   rename(`Number of Records` = N) |>
#   knitr::kable()

tbl = d_all |>
  # NA is Unknown:
  mutate(mode = case_when(
    is.na(mode) ~ "unknown",
    TRUE ~ mode
  )) |>
  count(mode, sort = TRUE) |>
  mutate(percent_of_records = n / nrow(d_all)) |>
  mutate(percent_of_records = round(percent_of_records, 3) * 100) |>
  # Reclassify Unknown to unknown:
  arrange(desc(n)) |>
  rename_all(snakecase::to_title_case) |>
  # Rename N to "Number of records"
  rename(`Number` = N, `Percent` = `Percent of Records`) |>
  gt() |>
  gt_plt_bar_pct(column = `Percent`, fill = "#252525", scaled = TRUE, labels=TRUE, width=120, height=30, font_size = "15px") |> 
  cols_width(1 ~ px(400), 2 ~ px(80)) |> 
  tab_style(
    style = cell_text(size=px(18)),
    locations = cells_body(columns = Mode)
  ) |>
  gt_theme_espn()
# tbl
# gt::gtsave(tbl, "tbl-mode.png")
knitr::include_graphics("tbl-mode.png")
```

@@

The cross-tabulation of reporting mode and offence type is shown in @tbl-mode-offences-crosstab, revealing some interesting tendencies.

@@

```{r}
#| label: tbl-mode-offences-crosstab
#| tbl-cap: Mode of transport from which video was recorded (columns) and offence type (rows), with counts and percentages within each reporting mode.
# pivot_combined |>
#   knitr::kable(digits = 1)
tbl = pivot_combined |>
  select(-`Total (%)`) |> 
  filter(offence_simple!="Total") |> 
  select(-c(Cyclist:Other)) |> 
  pivot_longer(cols=`Cyclist (%)`:`Other (%)`, names_to="mode", values_to="perc") |> 
  mutate(
    mode=factor(str_extract(mode, "^[^\ ]+"), levels=c("Driver", "Cyclist", "Other")),
    offence_simple=factor(offence_simple, levels=c("Inconsiderate driving", "Careless driving", "Other"))
    ) |> 
  group_by(offence_simple) |> 
  summarise(perc=list(round(perc)), count=first(Total)) |> 
  gt() |>
  gt_plt_bar_stack(perc, width=65, labels = c(" Cyclists ", " Drivers ", " Other "),
                   palette= c("#e31a1c", "#1f78b4", "#bdbdbd")) |>
    cols_width(1 ~ px(180), 2 ~ px(400), 3 ~ px(150)) |> 
  tab_style(
    style = cell_text(size=px(18)),
    locations = cells_body(columns = c(offence_simple, count))
    ) |> 
    gt_theme_espn() 
# gt::gtsave(tbl, "tbl-mode-offences-crosstab.png")
knitr::include_graphics("tbl-mode-offences-crosstab.png")
```

@@

Disposal categories assigned by police are shown in @tbl-disposal.
Roughly a third of cases (`r round(perc_no_further_action, 1)`%) resulted in no further action, and for most of the remainder drivers were required to undertake an education course (`r round(perc_course, 1)`%).
Conditional offers, that is, drivers being offered a reduced penalty for admitting guilt, were the third most common outcome.
Close to two percent of cases went to court and a further one percent underwent further investigation.

@@

There are `r unique(d_with_location$location) |> length()` unique locations (addresses) in the data, with the most common locations shown in @tbl-locations.

@@

There was an increase in submissions over time between 2021 and 2023.
This may reflect changing levels of road use during and after the Covid-19 pandemic, increased awareness of OpSnap, or increased video camera ownership levels.
Comparative analysis including other police jurisdictions will be able to determine whether this experience is common or isolated.
We hypothesise that it will prove to be a common experience and that that the volume of submissions will continue to increase in the near and middle term (though ideally, a decline would occur as roads become safer).

@@

Studies focused on specific types of road user will prove informative with respect to policy and practice.
Dash-cam submissions by motorists offers great potential to inform other aspects of road safety.
Research into submissions by pedestrians, horse riders and motorcyclists should prove viable to improve the safety of these parties.
There were over four hundred submissions by horse riders in West Yorkshire, which means there will be thousands nationwide.
The British Horse Society[^1] and others may be interested in this data being used to inform the safety of horses and riders.
The British Motorcyclists Federation[^2] may be interested in research promoting the safety of its members and in why motorcyclist submissions are relatively infrequent
.

>>>>

There were `r nrow(d_all) |> scales::number(big.mark = ",")` records in the dataset for the three-year study period, with a strong upward trend, as shown in the monthly counts presented in @fig-time.
Since early 2022, there have been more monthly records in the OpSnap data than in the official 'STATS19' road traffic collision records for West Yorkshire, highlighting the under-reporting of road traffic incidents in official statistics.
STATS19 records are from the Department for Transport's database of road traffic collisions reported to the police and only include incidents that result in injury.
Like OpSnap data, STATS19 records are open access.
For the results presented in @fig-time, STATS19 datasets were downloaded with the `stats19` R package [@lovelace2019] and filtered to include only records from West Yorkshire.

@@

```{r}
#| label: tbl-offences
#| include: true
#| tbl-cap: "Offence types reported (top 6 and other)."
#| width: 60%
# Aim: get table of n. offences by mode
d_mode_offence_count = d_all |>
  count(mode, offence, sort = TRUE) 
# offences in order of n. offences
d_offence_count = d_all |>
  count(offence, sort = TRUE)
# d_offence_count |>
#   arrange(desc(n)) |>
#   head(20) |>
#   knitr::kable()
# |offence                                                                                                |    n|
# |:------------------------------------------------------------------------------------------------------|----:|
# |n/a                                                                                                    | 5706|
# |rt88576 drive without reasonable consideration to others                                               | 4992|
# |rt88575 drive without due care and attention                                                           | 2917|
# |rt88975 drive motor vehicle fail to comply with red / green arrow / lane closure traffic light signals | 1364|
# |rt88971 fail to comply with red traffic light                                                          |  679|
# |rt88966 motor vehicle fail to comply with endorsable s36 traffic sign                                  |  411|
# |rv86019 use a handheld phone / device whilst driving a motor vehicle on a road                         |  357|
# |rt88760 fail to comply with solid white lines                                                          |  265|
# |rt88751 contravene give way sign                                                                       |  264|
# |suspected contravene weight restriction.                                                               |  213|
# |rt88751 contravene mandatory direction arrows                                                          |  212|
# |me82009 driving on hard shoulder of motorway                                                           |  113|
# |rt88975 fail to comply with red traffic light                                                          |  109|
# |rt88751 motor vehicle fail to comply with a non-endorsable traffic sign other (specify)                |   91|
# |zp97004 fail to comply with red light pelican crossing                                                 |   84|
# |zp97003 stop within controlled area of pelican crossing                                                |   81|
# |rc86814 driver not in proper control of vehicle                                                        |   62|
# |zp97001 stop vehicle within limits of pelican crossing                                                 |   59|
# |hy35001 drive/ride on footpath beside a road                                                           |   42|
# |rr84171 vehicle contravene local traffic order other than parking (e.g. bus lane)                      |   41|
# Pull out the top 6 offences excluding n/a:
d_offence_top_6 = d_mode_offence_count |>
  # filter(offence != "n/a") |>
  group_by(offence) |>
  summarise(n = sum(n)) |>
  arrange(desc(n)) |>
  head(6)
d_offence_classified = d_offence_count |>
  mutate(
    Offence = case_when(
      offence %in% d_offence_top_6$offence ~ offence,
      TRUE ~ "Other"
    )
  ) |>
  group_by(Offence) |>
  summarise(`Number of records` = sum(n)) |>
  # Arrange in descending order of n except for "Other":
  arrange(Offence == "Other", desc(`Number of records`)) |>
  # Rename "n/a" to "No offence":
  mutate(Offence = case_when(
    Offence == "n/a" ~ "NA No offence or unknown offence type",
    TRUE ~ Offence
  )) |>
  mutate(`Percent of records` = round(`Number of records` / sum(`Number of records`) * 100, 1))

d_offence_totals = d_offence_classified |>
  summarise(`Number of records` = sum(`Number of records`), `Percent of records` = sum(`Percent of records`)) |>
  mutate(Offence = "Total")

@@

tbl = d_offence_classified |>
  bind_rows(d_offence_totals) |>
  mutate(
    `Percent of records`= round(`Percent of records`)
    ) |> 
  rename(`Number` = `Number of records`, `Percent`=`Percent of records`) |> 
  gt() |>

  gt_plt_bar_pct(column = `Percent`, fill = "#252525", scaled = TRUE, labels=TRUE, width=120, height=30, font_size = "15px") |> 
  text_transform(
    fn <- function(x){
      code <- str_extract(x, "^[^\ ]+")
      desc <- str_remove(x, "^[^\ ]+")
      glue::glue("<em><span style='font-size:14px'>{code}</span></em><br><span style='font-size:18px'>{desc}</span>")
      },
    locations=cells_body(columns=Offence)
  ) |> 
  cols_width(1 ~ px(600), 2 ~ px(80)) |> 
  gt_theme_espn()
# gt::gtsave(tbl, "tbl-offences.html")
# webshot2::webshot("tbl-offences.html", "tbl-offences.png")
# browseURL("tbl-offences.png")
knitr::include_graphics("tbl-offences.png")
```

```{r}
# reclassify offences into types
x = d_all
regroup_offences_simple = function(x) {
 x |> 
   dplyr::mutate(
     offence_simple = dplyr::case_when(
       stringr::str_detect(offence, "drive without reasonable consideration") ~ "Inconsiderate driving",
       stringr::str_detect(offence, "drive without due care") ~ "Careless driving",
       TRUE ~ "Other"
     )
   )
}
# table(x$mode)
regroup_modes = function(x) {
  x |> 
    dplyr::mutate(
      mode_simplified = dplyr::case_when(
      mode == "cyclist" ~ "Cyclist",
      mode == "vehicle driver" ~ "Driver",
      TRUE ~ "Other"
    ))
}

@@

```{r}
#| label: tbl-mode
#| tbl-cap: Mode of transport of person submitting video evidence
#| width: "80%"
# d_all |>
#   count(mode, sort = TRUE) |>
#   mutate(percent_of_records = n / nrow(d_all)) |>
#   mutate(percent_of_records = round(percent_of_records, 3) * 100) |>
#   arrange(desc(n)) |>
#   rename_all(snakecase::to_title_case) |>
#   # Rename N to "Number of records"
#   rename(`Number of Records` = N) |>
#   knitr::kable()

tbl = d_all |>
  # NA is Unknown:
  mutate(mode = case_when(
    is.na(mode) ~ "unknown",
    TRUE ~ mode
  )) |>
  count(mode, sort = TRUE) |>
  mutate(percent_of_records = n / nrow(d_all)) |>
  mutate(percent_of_records = round(percent_of_records, 3) * 100) |>
  # Reclassify Unknown to unknown:
  arrange(desc(n))
tbl_totals = tbl |>
  summarise(n = sum(n), percent_of_records = sum(percent_of_records)) |>
  mutate(percent_of_records = round(percent_of_records)) |>
  mutate(mode = "Total")
tbl = bind_rows(tbl, tbl_totals) |>
  rename_all(snakecase::to_title_case) |>
  # Rename N to "Number of records"
  rename(`Number` = N, `Percent` = `Percent of Records`) |>
  gt() |>
  gt_plt_bar_pct(column = `Percent`, fill = "#252525", scaled = TRUE, labels=TRUE, width=120, height=30, font_size = "15px") |> 
  cols_width(1 ~ px(400), 2 ~ px(80)) |> 
  tab_style(
    style = cell_text(size=px(18)),
    locations = cells_body(columns = Mode)
  ) |>
  gt_theme_espn()
# tbl
# gt::gtsave(tbl, "tbl-mode.html")
# webshot2::webshot("tbl-mode.html", "tbl-mode.png")
knitr::include_graphics("tbl-mode.png")
```

@@

A cross-tabulation of transport mode (of the person submitting video evidence) and the offence type is shown in @tbl-mode-offences-crosstab.
For both variables it shows the two main categories plus an 'other' category.

@@

```{r}
#| label: tbl-mode-offences-crosstab
#| tbl-cap: Mode of transport of person submitting video (columns).
# pivot_combined |>
#   knitr::kable(digits = 1)
tbl = pivot_combined |>
  select(-`Total (%)`) |> 
  filter(offence_simple!="Total") |> 
  select(-c(Cyclist:Other)) |> 
  pivot_longer(cols=`Cyclist (%)`:`Other (%)`, names_to="mode", values_to="perc") |> 
  mutate(
    mode=factor(str_extract(mode, "^[^\ ]+"), levels=c("Driver", "Cyclist", "Other")),
    offence_simple=factor(offence_simple, levels=c("Inconsiderate driving", "Careless driving", "Other"))
    ) |> 
  group_by(offence_simple) |> 
  summarise(perc=list(round(perc)), count=first(Total)) |> 
  gt() |>
  gt_plt_bar_stack(perc, width=65, labels = c(" Cyclists ", " Drivers ", " Other "),
                   palette= c("#e31a1c", "#1f78b4", "#bdbdbd")) |>
    cols_width(1 ~ px(180), 2 ~ px(400), 3 ~ px(150)) |> 
  tab_style(
    style = cell_text(size=px(18)),
    locations = cells_body(columns = c(offence_simple, count))
    ) |> 
    gt_theme_espn() 
# gt::gtsave(tbl, "tbl-mode-offences-crosstab.png")
knitr::include_graphics("tbl-mode-offences-crosstab.png")
```

@@

Disposal categories assigned by police are shown in @tbl-disposal.
Roughly a third of cases (`r round(perc_no_further_action, 1)`%) resulted in no further action and, for most of the remainder drivers were required to undertake an education course.
Conditional offers, that is, drivers being offered a reduced penalty for admitting guilt, were the third most common outcome.
Nearly two percent of cases went to court and a further one percent underwent further investigation.

@@

There were `r unique(d_with_location$location) |> length()` unique locations (addresses) in the data, with the most common locations shown in @tbl-locations.

@@

There was an increase in submissions to opsnap over time between 2021 and 2023.
This may reflect changing levels of road use during and after the Covid-19 pandemic, increased awareness of OpSnap, or increased video camera ownership levels.
Comparative analysis including other police jurisdictions should be undertaken to determine whether this experience is common or isolated.
We hypothesise that it is a common experience and also that that the volume of submissions will continue to increase in the near and middle term (though ideally, a decline would occur as roads become safer).

@@

Studies focused on specific types of road user will prove informative with respect to policy and practice.
Dash-cam submissions by motorists offers great potential to inform other aspects of road safety.
Research into submissions by pedestrians, horse riders and motorcyclists should prove viable to improve the safety of these parties.
There were over four hundred submissions by horse riders in West Yorkshire, which means there will be thousands nationwide.
The British Horse Society[^1] and others may be interested in this data being used to inform the safety of horses and riders.
The British Motorcyclists Federation[^2] may be interested in research promoting the safety of its members and in why motorcyclist submissions are relatively infrequent.
